---
title: Identifying Priority Conservation Areas in the Florida Everglades for Mangrove Habitat
subtitle: Final Project, ENV 872
author: Isaac Benaka, Hanna Bliska, and Caroline Rowley
date: Fall 2022
output:
 bookdown::pdf_document2
 bookdown::html_document2
editor_options: 
  chunk_output_type: console
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
#setwd("~/Desktop/BenakaBliskaRowley_ENV872_EDA_FinalProject")
getwd()

# Load your packages
library(tidyverse)
library(lubridate)
library(scales)
library(cowplot)
library(trend)
library(zoo)
library(Kendall)
library(sf)
library(mapview)
library(agricolae)
library(BIOMASS)

# Set your ggplot theme
my_theme <- theme_classic(base_size=12) + theme(
  axis.text=element_text(color="black"), 
  legend.position="right")
theme_set(my_theme)

# Load your datasets
FCE.site.coordinates <- read.csv("./Data/Raw/FCE_LTER_site_coordinates.csv")
FCE.nutrients.data <- read.csv("./Data/Raw/FCE_LTER_Nutrients.csv",
                               stringsAsFactors = TRUE)
FCE.tree.data <- read.csv("./Data/Raw/FCE_LTER_Mangroves.csv",
                          stringsAsFactors = TRUE)
inat.birds <- read.csv("./Data/Raw/inaturalist_birds.csv",
                       stringsAsFactors = TRUE)
inat.reptiles <- read.csv("./Data/Raw/inaturalist_reptiles2.csv",
                          stringsAsFactors = TRUE)
inat.mammals <- read.csv("./Data/Raw/inaturalist_mammals.csv",
                         stringsAsFactors = TRUE)  
inat.fishnbugs <- read.csv("./Data/Raw/inaturalist_fish_arachninds_insects.csv", 
                           stringsAsFactors = TRUE)  
inat.amphibians <- read.csv("./Data/Raw/inaturalist_amphibians.csv",
                            stringsAsFactors = TRUE)
inat.plants <- read.csv("./Data/Raw/inaturalist_plants_fungi.csv",
                        stringsAsFactors = TRUE)
mangrove.sf <- st_read("./Data/Spatial/Mangrove_Habitat_in_Florida/Mangrove_Habitat_in_Florida.shp")
parks.boundary <- st_read("./Data/Spatial/park/nps_boundary.shp")

```


# Rationale and Research Questions

Mangroves are vital coastal ecosystems that support biological diversity and fortify fragile coastlines. Mangrove trees have adapted to live in water with high salinity, and thus these tropical forest trees have created nurseries and habitat for coastal species along sea-land interfaces (Duke et al., 2014). Further, mangrove ecosystems strengthen the shoreline, as they act as a natural barrier against wave- and storm-induced erosion (Duke et al., 2014). As a consequence of human development, agricultural expansion, and climate-change related factors, the distribution and nature of mangrove communities is changing (Duke et al., 2014).  

The Florida Coastal Everglades Long Term Ecological Research (FCE LTER) Site has been monitoring mangrove ecosystems in the Florida Everglades National Park. Their long-term monitoring programs have helped scientists track long-term changes in the mangrove communities for many years (Florida Coastal Everglades LTER, 2022). Their publicly available datasets report variables related to mangrove health, such as height, diameter, species, and nutrient availability. Additionally, citizen scientists have visited the Florida Everglades National Park and collected data on their observations of mangrove-associated species over time through iNaturalist. These databases pose an opportunity for us to investigate factors that may impact mangrove health and consider areas of the park where mangrove ecosystem conservation efforts may be focused to best protect these vital ecosystems.

Our research questions were:
1. Where are mangrove forests spatially located and how has the biomass of mangroves changed over time?
2. Are native, mangrove-associated species decreasing in the park and where are these species observed?
3. How has nutrient availability changed over time and where are the areas of highest concern (i.e., highest nutrient levels) for mangroves?

\newpage

# Dataset Information

```{r, renaming columns, include=FALSE}

#renaming column names and removing repeated header row
colnames(FCE.mean.sea.levels) <- c(
  "Station_ID", "Year", "Month", "MSL")
FCE.mean.sea.levels=FCE.mean.sea.levels[-1,]

colnames(FCE.nutrients.data) <- c(
  "SITE_NAME", "DATE", "TIME", "Salinity", "TN", "TP",
  "TOC", "NH4", "NandN", "NO2", "SRP", "DOC", "NO3")
FCE.nutrients.data=FCE.nutrients.data[-1,]

colnames(FCE.sawgrass.data) <- c(
  "SITE_NAME", "DATE", "Sawgrass_AG_Biomass", "Sawgrass_AG_Biomass_STD",
  "Sawgrass_Culm_Density", "Sawgrass_Culm_Density_STD") 
FCE.sawgrass.data=FCE.sawgrass.data[-1,]

colnames(FCE.site.coordinates) <- c(
  "SITE", "Latitude (Decimal Degrees)", "Longitude (Decimal Degrees)",
  "Latitude (Degrees, minutes)", "Longitude (Degrees, minutes)",
  "Northing (UTM)", "Easting (UTM)")
FCE.site.coordinates=FCE.site.coordinates[-1,]

colnames(FCE.tree.data) <- c(
  "SITE", "DATE", "Plot_ID", "Tree_TagNumber",
  "Species.Tree", "Tree_DBH", "Tree_Height")
FCE.tree.data=FCE.tree.data[-1,]

#wrangling the inaturalist data sets to have the same columns
inat.birds <- inat.birds %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

inat.reptiles <- inat.reptiles %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

inat.mammals <- inat.mammals %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

inat.fishnbugs <- inat.fishnbugs %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

inat.amphibians <- inat.amphibians %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

inat.plants <- inat.plants %>% 
  select(observed_on, latitude, longitude, scientific_name, common_name, iconic_taxon_name, taxon_id, taxon_order_name, taxon_genus_name)

#combining the inaturalist data sets into one data frame
inat.all <- rbind(inat.amphibians, inat.birds, inat.fishnbugs, inat.mammals, inat.plants, inat.reptiles)

inat.animals <-rbind(inat.birds, inat.amphibians, inat.mammals, inat.reptiles)

```

```{r, formatting data sets, include=FALSE}
#formatting date columns
FCE.nutrients.data$DATE <- as.Date(FCE.nutrients.data$DATE, format="%Y-%m-%d")
FCE.sawgrass.data$DATE <- as.Date(FCE.sawgrass.data$DATE, format="%Y-%m-%d")
FCE.tree.data$DATE <- as.Date(FCE.tree.data$DATE, format = "%Y-%m-%d")
inat.animals$observed_on <- as.Date(inat.animals$observed_on, format("%Y-%m-%d"))

#formatting numeric values
FCE.nutrients.data$TP <- as.numeric(FCE.nutrients.data$TP)
FCE.nutrients.data$TN <- as.numeric(FCE.nutrients.data$TN)
FCE.tree.data$Tree_DBH <- as.numeric(FCE.tree.data$Tree_DBH)

#wrangling site coordinates data frame to have matching sites
FCE.site.coordinates$SITE <- as.character(FCE.site.coordinates$SITE) 
#convert to character
FCE.site.coordinates$SITE <- gsub('\\s+', '', FCE.site.coordinates$SITE) 
#remove spaces from SITE codes to join to other datasets
FCE.site.coordinates$SITE <- as.factor(FCE.site.coordinates$SITE) 
#convert back to factor
```

```{r, changing site coordinates to spatial data frame, include=FALSE}
site.locations.sf <- st_as_sf(
  FCE.site.coordinates, coords=c(
    'Longitude (Decimal Degrees)', 
    'Latitude (Decimal Degrees)'),crs=4326)
#creating spatial data frame
#specifying Longitude as x and Latitude as y
#using crs 4326

mapview(site.locations.sf) #confirming locations of sites

```

## Mangrove Data Sets
Data on mangrove forest growth in the Everglades National Park were obtained from the Florida Coastal Everglades Long Term Ecological Research Site (FCE LTER). We downloaded the data on November 15, 2022 using the Environmental Data Initiative Data Portal (https://portal.edirepository.org/nis/home.jsp), where publicly available data for LTERs can be accessed. The data set selected was titled Mangrove Forest Growth from the Shark River Slough, Everglades National Park (FCE), South Florida, USA, January 1995 - ongoing. 

To create a spatial visualization of mangroves and sites in Everglades National Park, we used shapefile data titled Mangrove Habitat in Florida from the Florida Fish and Wildlife Conservation Commission to provide a basic visualization of where mangroves exist in Everglades National Park (https://geodata.myfwc.com/datasets/myfwc::mangrove-habitat-in-florida-1/about). We also used a shapefile of national park boundaries from the National Park Service (https://irma.nps.gov/DataStore/Reference/Profile/2224545?lnv=True).

The forest growth dataset contained dendrologic data for four long-term monitoring sites in the Everglades National Park. The dataset contained diameter-at-breast-height (DBH) and height measurements for four key mangrove speices — *Rhizophora mangle, Laguncularia racemosa, Avicennia germinans,* and *Conacarpus erectus*. Wrangling was performed to join geographic coordinate data to each plot to create a spatial dataframe. Site coordinate data were first transformed to a spatial data frame (`site.locations.sf`) before joining. Utilizing the BIOMASS r package, the above ground biomass of each sampled tree was calculated using the DBH measurements and geographic coordinates contained in this dataset. This data allowed us to estimate mangrove biomass density across the four sites and where mangrove habitat might be decreasing. We elected to remove one of the sites from analysis because it did not overlap with the nutrient data sites, leaving us with three sites: SRS4, SRS5, and SRS6.

To conduct analysis, we wrangled the data into three separate monitoring site dataframes to compare them against one another and determine site-by-site trends. Because the number of observations was not consistent across the sites over time, we elected to compare the average above ground biomass for individual trees at each site as opposed to summing the above ground biomass estimations together. To do this, we used the `group_by` and `summarise` functions in R to find the mean above ground biomass of trees at each site for each year that observations were available.

The final data structure containing above ground biomass estimations for measured mangroves at the three sites of interest in Everglades National Park (`treeAGB`) is as follows:
| Variable   | Units      | Range                   | Mean    |
| ---------- | ---------  | ----------------------- | ------- |
| SITE       | N/A        | 3 sites                 | N/A     |
| DATE       | YYYY-MM-DD | 1995-01-13 - 2019-10-16 | N/A     |
| Species.Tree| Abbrev.   | 4 species               | N/A     |
| Tree_DBH   | cm         | 1 - 835                 | 334.3   |
| Tree_Height| m          | -9999 - 8               | N/A     |
| meanWD     | kg/m^3     | 0.61 - 0.8977           | 0.817   |
| AGB        | metric tons| 0 - 1239.2250           | 324.991 |
| geometry   | x,y coords | NA                      | NA      |
| YEAR       | YYYY       | 1995 - 2019             | 2008    |

The final yearly datasets for mean above ground biomass of trees by site (`treesSRS4`,`treesSRS5`,`treesSRS6`) is as follows:
| Variable   | Units      | Range         | Mean    |
| ---------- | ---------  | ------------- | ------- |
| YEAR       | YYYY       | 1995 - 2019   | N/A     |
| meanAGB    | metric tons| 0 - 1239.2250 | 324.991 |

The final spatial dataset for Everglades National Park boundary (`park.boundary`) is as follows:
| Variable   | Units            | Range         | Mean    |
| ---------- | ---------------- | ------------- | ------- |
| UNIT_NAME  | N/A              | N/A           | N/A     |
| geometry   | x,y coordinates  | N/A           | N/A     |

The final spatial dataset for the mangrove habitat area in Everglades National Park (`mangrove.area`) is as follows:
| Variable   | Units            | Range         | Mean    |
| ---------- | ---------------- | ------------- | ------- |
| OBJECTID   | N/A              | N/A           | N/A     |
| SOURCEDATE | YYYY             | 1999 - 2016   | N/A     |
| geometry   | x,y coordinates  | N/A           | N/A     |

```{r, wrangling for wood density and AGB across the three sites of interest, include=FALSE}
#create dataframe for Rhizophora mangle and calculate wood density
R.trees <- st_as_sf(FCE.tree.data.sites, coords = c("Longitude (Decimal Degrees)","Latitude (Decimal Degrees)"), crs=4326) %>%
  filter(Species.Tree=="R") %>%
  mutate(getWoodDensity(genus = "Rhizophora", species = "mangle")) %>%
  select(DATE, SITE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry)

#add geographic coordinates
R.trees <- merge(x=R.trees, y=FCE.site.coordinates, by="SITE", all.x = TRUE) 
R.trees$`Latitude (Decimal Degrees)` <- as.numeric(R.trees$`Latitude (Decimal Degrees)`)
R.trees$`Longitude (Decimal Degrees)` <- as.numeric(R.trees$`Longitude (Decimal Degrees)`)
  
#create dataframe of coordinates for each mangrove observation
R.trees.coordinates <- cbind(R.trees$`Longitude (Decimal Degrees)`, R.trees$`Latitude (Decimal Degrees)`)

#compute the above ground biomass in metric tons and add to species dataframe
R.trees <- R.trees %>%
  mutate(computeAGB(R.trees$Tree_DBH, R.trees$meanWD, coord = R.trees.coordinates)) %>% 
  rename(AGB = `computeAGB(R.trees$Tree_DBH, R.trees$meanWD, coord = R.trees.coordinates)`)
R.trees <- R.trees %>%
  select(SITE, DATE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry, AGB)

#create dataframe for Laguncularia racemosa and calculate wood density
L.trees <- st_as_sf(FCE.tree.data.sites, coords = c("Longitude (Decimal Degrees)","Latitude (Decimal Degrees)"), crs=4326) %>%
  filter(Species.Tree=="L") %>%
  mutate(getWoodDensity(genus = "Laguncularia", species = "racemosa")) %>%
  select(DATE, SITE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry)

#add geographic coordinates
L.trees <- merge(x=L.trees, y=FCE.site.coordinates, by="SITE", all.x = TRUE)
L.trees$`Latitude (Decimal Degrees)` <- as.numeric(L.trees$`Latitude (Decimal Degrees)`)
L.trees$`Longitude (Decimal Degrees)` <- as.numeric(L.trees$`Longitude (Decimal Degrees)`)
  
#create dataframe of coordinates for each mangrove observation
L.trees.coordinates <- cbind(L.trees$`Longitude (Decimal Degrees)`, L.trees$`Latitude (Decimal Degrees)`)

#compute the above ground biomass in metric tons and add to species dataframe
L.trees <- L.trees %>%
  mutate(computeAGB(L.trees$Tree_DBH, L.trees$meanWD, coord = L.trees.coordinates)) %>% 
  rename(AGB = `computeAGB(L.trees$Tree_DBH, L.trees$meanWD, coord = L.trees.coordinates)`)
L.trees <- L.trees %>%
  select(SITE, DATE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry, AGB)

#create dataframe for Avicennia germinans and calculate wood density
A.trees <- st_as_sf(FCE.tree.data.sites, coords = c("Longitude (Decimal Degrees)","Latitude (Decimal Degrees)"), crs=4326) %>%
  filter(Species.Tree=="A") %>%
  mutate(getWoodDensity(genus = "Avicennia", species = "germinans")) %>%
  select(DATE, SITE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry)

#add geographic coordinates
A.trees <- merge(x=A.trees, y=FCE.site.coordinates, by="SITE", all.x = TRUE)
A.trees$`Latitude (Decimal Degrees)` <- as.numeric(A.trees$`Latitude (Decimal Degrees)`)
A.trees$`Longitude (Decimal Degrees)` <- as.numeric(A.trees$`Longitude (Decimal Degrees)`)

#create dataframe of coordinates for each mangrove observation
A.trees.coordinates <- cbind(A.trees$`Longitude (Decimal Degrees)`, A.trees$`Latitude (Decimal Degrees)`)

#compute the above ground biomass in metric tons and add to species dataframe
A.trees <- A.trees %>%
  mutate(computeAGB(A.trees$Tree_DBH, A.trees$meanWD, coord = A.trees.coordinates)) %>% 
  rename(AGB = `computeAGB(A.trees$Tree_DBH, A.trees$meanWD, coord = A.trees.coordinates)`)
A.trees <- A.trees %>%
  select(SITE, DATE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry, AGB)

#create dataframe for Conacarpus erectus and calculate wood density
C.trees <- st_as_sf(FCE.tree.data.sites, coords = c("Longitude (Decimal Degrees)","Latitude (Decimal Degrees)"), crs=4326) %>%
  filter(Species.Tree=="C") %>%
  mutate(meanWD = rep(0.69)) %>% #wood density value for Conacarpus erectus
  select(DATE, SITE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry)

#add geographic coordinates
C.trees <- merge(x=C.trees, y=FCE.site.coordinates, by="SITE", all.x = TRUE)
C.trees$`Latitude (Decimal Degrees)` <- as.numeric(C.trees$`Latitude (Decimal Degrees)`)
C.trees$`Longitude (Decimal Degrees)` <- as.numeric(C.trees$`Longitude (Decimal Degrees)`)

#create dataframe of coordinates for each mangrove observation
C.trees.coordinates <- cbind(C.trees$`Longitude (Decimal Degrees)`, C.trees$`Latitude (Decimal Degrees)`)

#compute the above ground biomass in metric tons and add to species dataframe
C.trees <- C.trees %>%
  mutate(computeAGB(C.trees$Tree_DBH, C.trees$meanWD, coord = C.trees.coordinates)) %>% 
  rename(AGB = `computeAGB(C.trees$Tree_DBH, C.trees$meanWD, coord = C.trees.coordinates)`)
C.trees <- C.trees %>%
  select(SITE, DATE, Species.Tree, Tree_DBH, Tree_Height, meanWD, geometry, AGB)

treeAGB <- bind_rows(A.trees, C.trees, L.trees, R.trees) %>% #combine species dataframes together
  filter(SITE != "TS/Ph8") %>% #remove site outside of area of interest
  mutate(YEAR = year(DATE)) #change YEAR column values to date class

#saving as a processed data frame
write.csv(treeAGB, row.names = FALSE, file= "./Data/Processed/treeAGB.csv")

```

```{r, wrangling into site dataframes, include=FALSE}
Years <- as.data.frame(seq(2000,2019,1)) #create dataframe with list of years
colnames(Years) <- c("YEAR") #add complete list of years to dataframes that are missing observations for those years


treesSRS4 <- treeAGB %>%
  filter(SITE == "SRS4") %>% #select for site
  group_by(YEAR) %>% #grouping by year
  summarise(meanAGB = mean(AGB)) #find mean AGB for each year
treesSRS4 <- left_join(Years, treesSRS4, by = c("YEAR")) #add years where data is missing for interpolation
#saving as a processed data frame
write.csv(treesSRS4, row.names = FALSE, file= "./Data/Processed/treesSRS4.csv")


treesSRS5 <- treeAGB %>%
  filter(SITE == "SRS5") %>% #select for site
  group_by(YEAR) %>% #grouping by year
  summarise(meanAGB = mean(AGB)) #find mean AGB for each year
treesSRS5 <- left_join(Years, treesSRS5, by = c("YEAR"))#add years where data is missing for interpolation
#saving as a processed data frame
write.csv(treesSRS5, row.names = FALSE, file= "./Data/Processed/treesSRS5.csv")


treesSRS6 <- treeAGB %>%
  filter(SITE == "SRS6") %>% #select for site
  group_by(YEAR) %>% #grouping by year
  summarise(meanAGB = mean(AGB)) #find mean AGB for each year
treesSRS6 <- left_join(Years, treesSRS6, by = c("YEAR"))#add years where data is missing for interpolation
#saving as a processed data frame
write.csv(treesSRS6, row.names = FALSE, file= "./Data/Processed/treesSRS6.csv")

```

```{r, wrangling for Everglades National Park boundary polygon, include=FALSE}
#Filter for Everglades National Park boundary
park.boundary <- parks.boundary %>% filter(UNIT_NAME == "Everglades National Park") %>% select(UNIT_NAME, geometry) #pare down columns

# Project spatial data to UTM 17N
park.utm <- st_transform(park.boundary,32617)

```

```{r, wrangling mangrove habtiat shapefile, include=FALSE}
# Project spatial data to UTM 17N
mangrove.utm <- st_transform(mangrove.sf,32617)

# Crop mangrove shapefile to Everglades National Park boundary.
mangrove.area <- mangrove.utm[park.utm,] %>% group_by(SOURCEDATE) %>% select(OBJECTID, SOURCEDATE, geometry)

```


## Species Data Set

## Nutrient Data Set
Data on nutrient concentrations in the Everglades National Park were obtained from the Florida Coastal Everglades Long Term Ecological Research Site (FCE LTER). We downloaded the data on November 15, 2022 using the Environmental Data Initiative Data Portal (https://portal.edirepository.org/nis/home.jsp), where publicly available data for LTERs can be accessed. The data set selected was titled Water Quality Data (Grab Samples) from the Shark River Slough, Everglades National Park (FCE LTER), May 2001 - ongoing.  

This data set obtained nutrient data for 8 long-term mangrove vegetation monitoring plots. Since mangroves are nutrient-sensitive species (Feller et al. 2007), this data set would allow us to understand where mangrove ecosystems may be stressed due to nutrient concentrations. The data set was initially wrangled for utility in R; we renamed columns and changed the format of the data. Due to the sensitivity of mangroves to nitrogen and phosphorus (Feller et al. 2007), we chose to focus our analyses on these two nutrients. Additional wrangling was then performed to join the coordinate data to each plot to create a spatial data frame. Site coordinate data were first transformed to a spatial data frame (`site.locations.sf`) before joining.

For time series analysis, we wrangled the data to determine the mean monthly and yearly nitrogen and phosphorus measurements within the monitoring sites. For the yearly measurements, we extracted the years from the date observations to create a new year column, and then averaged the nitrogen and phosphorus measurements within each year using the `group_by` and `summarize` functions in R. For the monthly measurements, we extracted the months from the date observations to create a new month column, and then averaged the nitrogen and phosphorus measurements within each month using the `group_by` and `summarize` functions in R. We then created a sequence of all of the months between the first and last data observation (May 2001 to December 2021) and joined this data to the original data frame to ensure there was an observation for every month between the first and last data observation date.

The final data structure for the yearly data set for phosphorus (`FCE.phos.yearly.data`) is as follows:
| Variable   | Units      | Range         | Mean    |
| ---------- | ---------  | ------------- | ------- |
| YEAR       | YYYY       | 2001-2021     | N/A     |
| MEAN_TP    | mcmol/L    | 21.64 - 75.26 | 45.36   |


The final data structure for the yearly data set for nitrogen (`FCE.nitro.yearly.data`) is as follows:
| Variable   | Units      | Range         | Mean    |
| ---------- | ---------  | ------------- | ------- |
| YEAR       | YYYY       | 2001-2021     | N/A     |
| MEAN_TN    | mcmol/L    | 416.7 - 950.3 | 699.2   |

The final data structure for the monthly data set for phosphorus (`FCE.phos.monthly.data`) is as follows:
| Variable   | Units      | Range                   | Mean    |
| ---------- | ---------  | ----------------------- | ------- |
| Date       | YYYY-MM-DD | 2001-05-01 - 2021-12-01 | N/A     |
| year       | YYYY       | 2001-2021               | N/A     |
| month      | MM         | 1-12                    | N/A     |
| MEAN_TP    | mcmol/L    | 5.167-141.833           | 46.021  |

The final data structure for the monthly data set for nitrogen (`FCE.nitro.monthly.data`) is as follows:
| Variable   | Units      | Range                   | Mean    |
| ---------- | ---------  | ----------------------- | ------- |
| Date       | YYYY-MM-DD | 2001-05-01 - 2021-12-01 | N/A     |
| year       | YYYY       | 2001-2021               | N/A     |
| month      | MM         | 1-12                    | N/A     |
| MEAN_TN    | mcmol/L    | 147-1232                | 698.3   |

For site comparison analysis, we wrangled the data to select only nitrogen and phosphorus for each site using the `select` function in R. 

The final data structure for the data set (`FCE.site.nutrient.data`) is as follows:
| Variable   | Units            | Range                   | Mean    |
| ---------- | ---------------- | ----------------------- | ------- |
| SITE_NAME  | N/A              | 8 different sites       | N/A     |
| DATE       | YYYY-mm-dd       | 2001-05-23 - 2021-12-06 | N/A     |
| TP         | mcmol/L          | 1 - 155                 | 45.6    |
| TN         | mcmol/L          | 1 - 1394                | 698.6   |

For spatial visualizations of the site nutrient data, we created a spatial data frame using the site coordinate data.

The final data structure for the data set (`site.locations.nutrients.sf`) is as follows:
| Variable   | Units            | Range                   | Mean    |
| ---------- | ---------------- | ----------------------- | ------- |
| SITE_NAME  | N/A              | 8 different sites       | N/A     |
| DATE       | YYYY-mm-dd       | 2001-05-23 - 2021-12-06 | N/A     |
| TP         | mcmol/L          | 1 - 155                 | 45.6    |
| TN         | mcmol/L          | 1 - 1394                | 698.6   |
| Latitude   | Degrees, minutes | N/A                     | N/A     |
| Longitude  | Degrees, minutes | N/A                     | N/A     |
| Northing   | UTM degrees N    | N/A                     | N/A     |
| Easting    | UTM degrees W    | N/A                     | N/A     |
| Geometry   | x,y coordinates  | N/A                     | N/A     |

```{r, wrangling nutrients data set for mean P and N for each year, include=FALSE}
FCE.phos.yearly.data <- FCE.nutrients.data %>%
  select(SITE_NAME, DATE, TP) %>% #selecting variables of interest
  mutate(YEAR=year(DATE)) %>% #creating year column
  group_by(YEAR) %>% #grouping by year
  summarise(MEAN_TP=mean(TP)) #averaging the TP for each year

#saving as processed data frame
write.csv(FCE.phos.yearly.data, row.names = FALSE, file= "./Data/Processed/FCE.phos.yearly.data.csv")
  
FCE.nitro.yearly.data <- FCE.nutrients.data %>%
  select(SITE_NAME, DATE, TN) %>% #selecting variables of interest
  mutate(YEAR=year(DATE)) %>% #creating year column
  group_by(YEAR) %>% #grouping by year
  summarise(MEAN_TN=mean(TN))  #averaging the TN for each year

#saving as a processed data frame
write.csv(FCE.nitro.yearly.data, row.names = FALSE, file= "./Data/Processed/FCE.nitro.yearly.data.csv")
```

```{r, wranging nutrients data for mean P and N for each month, include=FALSE}
#monthly phosphorus averages
FCE.phos.monthly.data <- FCE.nutrients.data %>%
  select(SITE_NAME, DATE, TP) %>%
  mutate(month = month(DATE)) %>% #adding month column
  mutate(year = year(DATE)) %>% #adding a year column
  group_by(year, month) %>% #grouping by year and month
  summarise(MEAN_TP=mean(TP)) #averaging the TP for each month of the year

FCE.phos.monthly.data <- FCE.phos.monthly.data %>%
  mutate("Date"=my(paste0(month,"-",year))) #creating a new date column
  #observations are for the first of the month

Months <- as.data.frame(seq(
  as.Date("2001-05-01"), as.Date("2021-12-01"),"months")) 
#creating a data frame of monthly dates from May 2001 to December 2021
#will ultimately allow us to identify months that do not have observations
#and then interpolate those at a later point

colnames(Months) <- c("Date") #renaming column name

FCE.phos.monthly.data <- left_join(
  Months, FCE.phos.monthly.data, by = c("Date"))
#joining data frames so that there is an observation for 
#each month of the year from May 2001 to December 2021

#saving as a processed data frame
write.csv(FCE.phos.monthly.data, row.names = FALSE, file= "./Data/Processed/FCE.phos.monthly.data.csv")
  
#nitrogen 
FCE.nitro.monthly.data <- FCE.nutrients.data %>%
  select(SITE_NAME, DATE, TN) %>%
  mutate(month = month(DATE)) %>% #adding month column
  mutate(year = year(DATE)) %>% #adding a year column
  group_by(year, month) %>% #grouping by year and month
  summarise(MEAN_TN=mean(TN)) #averaging the TN for each month of the year

FCE.nitro.monthly.data <- FCE.nitro.monthly.data %>%
  mutate("Date"=my(paste0(month,"-",year))) #creating a new date column
  #observations are for the first of the month

FCE.nitro.monthly.data <- left_join(
  Months, FCE.nitro.monthly.data, by = c("Date"))
#joining data frames so that there is an observation for 
#each month of the year from May 2001 to December 2021

#saving as a processed data frame
write.csv(FCE.nitro.monthly.data, row.names = FALSE, file= "./Data/Processed/FCE.phos.monthly.data.csv")
```

```{r, wrangling nutrients data set for site differences, include=FALSE}
FCE.site.nutrient.data <- FCE.nutrients.data %>%
  select(SITE_NAME, DATE, TP, TN) #selecting variables 
#of interest for site comparisons

#saving as a processed data frame
write.csv(FCE.site.nutrient.data, row.names = FALSE, file= "./Data/Processed/FCE.phos.monthly.data.csv")

```

```{r, adding site coordinates to the nutrient data set, include=FALSE}
site.locations.nutrients.sf <-
  merge(x=site.locations.sf, 
        y=FCE.site.nutrient.data,
        by.x="SITE",
        by.y="SITE_NAME")
#joining the site locations spatial data frames
#to the sites with nutrient data using merge

```

\newpage

# Exploratory Analysis 

## Mangrove Data

To address Research Question #1, we sought to explore trends in above ground biomass for mangroves in the Florida Everglades National Park over time. We focused on three mangrove monitoring sites that overlapped with nutrient data collection sites (See Figure \@ref(fig:mangrovemap)). The three sites displayed a wide range of above ground biomass for trees that were sampled from 1996 - present (See Figure \@ref(fig:mangrove.site.plot)). We calculated the mean above ground biomass for trees in each site across each year of data collection. Visualizing these data suggested that SRS4 displayed no trend in mean tree AGB growth, while SRS5 and SRS6 displayed possible increasing and decreasing trends respectively (See Figure \@ref(fig:AGB.site.plot)). However, to understand if the different sites had scientifically significant trends, we would need to conduct further analysis.

```{r mangrove distribution, fig.cap = "Map of Mangrove Monitoring Sites in Florida Coastal Everglades National Park", echo=FALSE, include=FALSE}
# Generate map of mangrove habitat in Everglades National Park.
mangrovemap <- mapview(mangrove.area, color = "green", col.regions = "dark green", alpha.regions = .4, layer.name = "Mangrove Habitat") + mapview(park.utm, color = 'black', lwd = 2, col.regions = "black", alpha.regions = 0, layer.name = "Everglades National Park Boundary")+
mapview(treeAGB, layer.name = "Monitoring Sites")
mangrovemap
```

```{r mangrove.site.plot, fig.cap = "Above Ground Biomass of Mangrove Trees by Site in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
mangrove.site.plot <- ggplot(treeAGB, aes(x=SITE, y=AGB)) +
  geom_boxplot()+
  xlab(expression("Site Name")) +
  ylab(expression("Above Ground Biomass (metric tons)")) +
  ggtitle(expression("Above Ground Biomass of Mangrove Trees by Site"),
          subtitle="Florida Coastal Everglades")
mangrove.site.plot
```

```{r AGB.sites.plot, fig.cap = "Mean Above Ground Biomass of Mangrove Trees by Year at Sites in Florida Everglades National Park", echo=FALSE, include=FALSE}
AGB.SRS4.plot <- ggplot(treesSRS4, aes(x=YEAR,y=meanAGB))+
  geom_point()+  
  geom_smooth(method = "lm", se=F)+
  labs(title="SRS4", y="Mean AGB (mt)")

AGB.SRS5.plot <- ggplot(treesSRS5, aes(x=YEAR,y=meanAGB))+
  geom_point()+  
  geom_smooth(method = "lm", se=F)+
  labs(title="SRS5", y="Mean AGB (mt)")

AGB.SRS6.plot <- ggplot(treesSRS6, aes(x=YEAR,y=meanAGB))+
  geom_point()+  
  geom_smooth(method = "lm", se=F)+
  labs(title="SRS6", y="Mean AGB (mt)")

AGB.sites.plot <- plot_grid(AGB.SRS4.plot, AGB.SRS5.plot, AGB.SRS6.plot, ncol=1)
AGB.sites.plot
```

## Species Data

## Nutrient Data

To address Research Question #3, we were interested in exploring trends of nutrient concentrations in the vegetation plots over time. We calculated mean measurements of phosphorus and nitrogen for each year in the Florida Coastal Everglades. Our initial visualization of the relationship between time and mean phosphorus as well as mean nitrogen suggested that there was likely an increasing trend in phosphorus concentrations over time (See Figure \@ref(fig:phos.plot)) and nitrogen concentrations over time (See Figure \@ref(fig:nitro.plot)). The results of these visualizations motivated us to later analyze if there was a significant increasing trend in nutrient concentrations over time.

```{r phos.plot, fig.cap = "Mean total phosphorus measurements over time in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
phos.plot <- ggplot(FCE.phos.yearly.data, aes(x=YEAR, y=MEAN_TP)) +
  geom_point()+
  geom_smooth(method=lm, se=FALSE, color="blue")+
  xlab(expression("Date")) +
  ylab(expression("Mean Total Phosphorus (mcmol/L)")) +
  ggtitle(expression("Mean Total Phosphorus Measurements"),
          subtitle="Florida Coastal Everglades")
phos.plot
```


```{r nitro.plot, fig.cap = "Mean total nitrogen measurements over time in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
nitro.plot <- ggplot(FCE.nitro.yearly.data, aes(x=YEAR, y=MEAN_TN)) +
  geom_point()+
  geom_smooth(method=lm, se=FALSE, color="blue") +
  xlab(expression("Date")) +
  ylab(expression("Mean Total Nitrogen (mcmol/L)")) +
  ggtitle(expression("Mean Total Nitrogen Measurements"),
          subtitle="Florida Coastal Everglades")
nitro.plot
```

We were also interested in identifying areas of mangrove habitat within the Florida Coastal Everglades that may be stressed due to high nutrient concentrations. Therefore, we initially examined if different mangrove vegetation plots in the park had different phosphorus and nitrogen concentrations. Our initial visualization suggested that some sites, such as site SRS6, may have higher phosphorus concentrations ((See Figure \@ref(fig:phos.site.plot)), See Figure \@ref(fig:map.site.phos.plot)). Similarly, our initial visualization suggested the site SRS1d may have higher nitrogen concentrations ((See Figure \@ref(fig:nitro.site.plot)), See Figure \@ref(fig:map.site.nitro.plot)). These visualizations motivated further analysis to determine if the concentrations at different sites were significantly different from one another.

```{r phos.site.plot, fig.cap = "Total phosphorus measurements for sites in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
phos.site.plot <- ggplot(FCE.site.nutrient.data, aes(x=SITE_NAME, y=TP)) +
  geom_boxplot()+
  xlab(expression("Site Name")) +
  ylab(expression("Total Phosphorus (mcmol/L)")) +
  ggtitle(expression("Total Phosphorus Measurements for Sites"),
          subtitle="Florida Coastal Everglades")
phos.site.plot
```


```{r map.site.phos.plot, fig.cap = "Map of Florida Coastal Everglades indicating level of phosphorus measurements per site", echo=FALSE, include=FALSE}
map.site.phos.plot <- mapview(site.locations.nutrients.sf, zcol="TP", at = seq(0, 60, 10))
map.site.phos.plot
```


```{r nitro.site.plot, fig.cap = "Total nitrogen measurements for sites in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
nitro.site.plot <- ggplot(FCE.site.nutrient.data, aes(x=SITE_NAME, y=TN)) +
  geom_boxplot()+
  xlab(expression("Site Name")) +
  ylab(expression("Total Nitrogen (mcmol/L)")) +
  ggtitle(expression("Total Nitrogen Measurements for Sites"),
          subtitle="Florida Coastal Everglades")
nitro.site.plot
```


```{r map.site.nitro.plot, fig.cap = "Map of Florida Coastal Everglades indicating level of nitrogen measurements per site", echo=FALSE, include=FALSE}
map.site.nitro.plot <- mapview(site.locations.nutrients.sf, zcol="TN")
map.site.nitro.plot
```

\newpage

# Analysis

## Question 1: Where are mangrove forests spatially located and how has the biomass of mangroves changed over time?

There was a significant increasing trend in mean monthly phosphorus levels in the Florida Coastal Everglades over time (Mann-Kendall test, score=12204, p-value =< 2.2e-16). It is evident that the trend has increasing at a relatively consistent rate over time (See Figure \@ref(fig:plot.mean.monthly.phos)). The Mann-Kendall test was performed after determining there was no seasonality in the data.

```{r, time series analysis for mangrove sites, include=FALSE}
treesSRS4$meanAGB <- na.approx(treesSRS4$meanAGB) #using`na.approx` to fill in missing yearly data

FCE.phos.monthly.ts <- ts(
  FCE.phos.monthly.data$MEAN_TP, 
  start=c(2001,5), end=c(2021,12),frequency=12) 
#generating time series object
#monthly measurements, using 12

FCE.phos.monthly.ts.decomposed <- stl(
  FCE.phos.monthly.ts, s.window = "periodic") 
#decomposing daily time series
plot(FCE.phos.monthly.ts.decomposed) #plotting components

#running mann kendall
FCE.phos.monthly.ts.trend <- MannKendall(FCE.phos.monthly.ts) 
#using `MannKendall` for trend analysis
summary(FCE.phos.monthly.ts.trend) #summary
```

```{r plot.mean.monthly.phos, fig.cap="Mean monthly phosphorus levels from 2001-2021 in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
plot.mean.monthly.phos <- ggplot(
  FCE.phos.monthly.data, aes(
    x=Date, y=MEAN_TP)) + #setting axes
  geom_point() + #generating scatter plot
  geom_line() + #generating line plot
  scale_x_date(date_breaks="1 year", labels=date_format("%m-%Y")) +
  #using `scale_x_date` to generate a break every 6 months
  #labels to show month-date-year
  theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  #angle for the day-month-year labels
  xlab(expression("Date")) + #setting x axis label
  ylab(expression ("Mean Monthly Phosphorus Levels (mcmol/L)")) +
#setting y axis label
  geom_smooth(method=lm, se = FALSE, color="red") + #adding linear trend
  ggtitle(expression("Mean Monthly Phosphorus Levels from 2001-2021"),
          subtitle="Florida Coastal Everglades")
print(plot.mean.monthly.phos) #printing plot
```
## Question 2: Are native, mangrove-associated species decreasing in the park and where are these species observed?

## Question 3: How has nutrient availability changed over time and where are the areas of highest concern for mangroves?

There was a significant increasing trend in mean monthly phosphorus levels in the Florida Coastal Everglades over time (Mann-Kendall test, score=12204, p-value =< 2.2e-16). It is evident that the trend has increasing at a relatively consistent rate over time (See Figure \@ref(fig:plot.mean.monthly.phos)). The Mann-Kendall test was performed after determining there was no seasonality in the data.

```{r, time series analysis for phosphorus, include=FALSE}
FCE.phos.monthly.data$MEAN_TP <- 
  na.approx(FCE.phos.monthly.data$MEAN_TP) #using`na.approx`
#to fill in missing monthly data

FCE.phos.monthly.ts <- ts(
  FCE.phos.monthly.data$MEAN_TP, 
  start=c(2001,5), end=c(2021,12),frequency=12) 
#generating time series object
#monthly measurements, using 12

FCE.phos.monthly.ts.decomposed <- stl(
  FCE.phos.monthly.ts, s.window = "periodic") 
#decomposing daily time series
plot(FCE.phos.monthly.ts.decomposed) #plotting components

#running mann kendall
FCE.phos.monthly.ts.trend <- MannKendall(FCE.phos.monthly.ts) 
#using `MannKendall` for trend analysis
summary(FCE.phos.monthly.ts.trend) #summary
```

```{r plot.mean.monthly.phos, fig.cap="Mean monthly phosphorus levels from 2001-2021 in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
plot.mean.monthly.phos <- ggplot(
  FCE.phos.monthly.data, aes(
    x=Date, y=MEAN_TP)) + #setting axes
  geom_point() + #generating scatter plot
  geom_line() + #generating line plot
  scale_x_date(date_breaks="1 year", labels=date_format("%m-%Y")) +
  #using `scale_x_date` to generate a break every 6 months
  #labels to show month-date-year
  theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  #angle for the day-month-year labels
  xlab(expression("Date")) + #setting x axis label
  ylab(expression ("Mean Monthly Phosphorus Levels (mcmol/L)")) +
#setting y axis label
  geom_smooth(method=lm, se = FALSE, color="red") + #adding linear trend
  ggtitle(expression("Mean Monthly Phosphorus Levels from 2001-2021"),
          subtitle="Florida Coastal Everglades")
print(plot.mean.monthly.phos) #printing plot
```

There was also significant increasing trend in mean monthly nitrogen levels in the Florida Coastal Everglades over time (Mann-Kendall test, score=9896, p-value =< 2.2e-16). Since 2001, mean monthly nitrogen levels have risen to over 750 mcmol/L (See Figure \@ref(fig:plot.mean.monthly.nitro)). 

```{r, time series analysis for nitrogen, include=FALSE}
FCE.nitro.monthly.data$MEAN_TN <- 
  na.approx(FCE.nitro.monthly.data$MEAN_TN) #using`na.approx`
#to fill in missing monthly data

FCE.nitro.monthly.ts <- ts(
  FCE.nitro.monthly.data$MEAN_TN, 
  start=c(2001,5), end=c(2021,12),frequency=12) 
#generating time series object
#monthly measurements, using 12

FCE.nitro.monthly.ts.decomposed <- stl(
  FCE.nitro.monthly.ts, s.window = "periodic") 
#decomposing daily time series
plot(FCE.nitro.monthly.ts.decomposed) #plotting components

#running mann kendall
FCE.nitro.monthly.ts.trend <- MannKendall(FCE.nitro.monthly.ts) 
#using `MannKendall` for trend analysis
summary(FCE.nitro.monthly.ts.trend) #summary
```

```{r plot.mean.monthly.nitro, fig.cap="Mean monthly nitrogen levels from 2001-2021 in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
plot.mean.monthly.nitro <- ggplot(
  FCE.nitro.monthly.data, aes(
    x=Date, y=MEAN_TN)) + #setting axes
  geom_point() + #generating scatter plot
  geom_line() + #generating line plot
  scale_x_date(date_breaks="1 year", labels=date_format("%m-%Y")) +
  #using `scale_x_date` to generate a break every 6 months
  #labels to show month-date-year
  theme(axis.text.x=element_text(angle=60, hjust=1)) + 
  #angle for the day-month-year labels
  xlab(expression("Date")) + #setting x axis label
  ylab(expression ("Mean Monthly Nitrogen Levels (mcmol/L)")) +
#setting y axis label
  geom_smooth(method=lm, se = FALSE, color="red") + #adding linear trend
  ggtitle(expression("Mean Monthly Nitrogen Levels from 2001-2021"),
          subtitle="Florida Coastal Everglades")
print(plot.mean.monthly.nitro) #printing plot
```

The results of the time series analyses indicate that mean monthly concentrations of nitrogen and phosphorus have increased over time in the Florida Coastal Everglades.

In the Florida Coastal Everglades, phosphorus concentrations varied significantly among the 8 mangrove habitat sites (ANOVA, df=7, F=28.75, p<0.001). The HSD test determined that site SRS6 had a mean phosphorus measurement that was statistically distinct from all other sites. SRS5 also had a mean phosphorus measurement that was statistically distinct from all other sites. It is evident that site SRS6 had the highest mean phosphorus measurements of all sites (See Figure \@ref(fig:anova.phos.site.plot)). 

```{r, site analysis for phosphorus, include=FALSE}
#running anova to see if particular sites have more phosphorus
phos.site.anova <- aov(data=FCE.site.nutrient.data, 
                           TP ~ SITE_NAME)
summary(phos.site.anova)

#evaluating groups
phos.site.anova.groups <- HSD.test(phos.site.anova, 
                                       "SITE_NAME", 
                                       group = TRUE)
#printing groups
phos.site.anova.groups

#results indicate that there are four groups of sites: a, b, c, d. There is only one site that is in group a and only one site in group b. 
```

```{r anova.phos.site.plot, fig.cap="Phosphorus measurements for mangrove habitat sites in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
anova.phos.site.plot <- ggplot(FCE.site.nutrient.data, aes(x=SITE_NAME, y=TP)) +
  geom_boxplot()+
  xlab(expression("Site Name")) +
  ylab(expression("Phosphorus (mcmol/L)")) +
  ggtitle(expression("Phosphorus Measurements for Sites"),
          subtitle="Florida Coastal Everglades") +
  annotate("text", x = 1, y = 200, label = "d") +
  annotate("text", x = 2, y = 200, label = "d") +
  annotate("text", x = 3, y = 200, label = "cd") +
  annotate("text", x = 4, y = 200, label = "d") +
  annotate("text", x = 5, y = 200, label = "cd") +
  annotate("text", x = 6, y = 200, label = "c") +
  annotate("text", x = 7, y = 200, label = "b") +
  annotate("text", x = 8, y = 200, label = "a")
anova.phos.site.plot
```

Nitrogen concentrations also varied significantly among the 8 mangrove habitat sites (ANOVA, df=7, F=18.68, p<0.001). The HSD test determined that the mean nitrogen concentrations of site SRS1d, SRS2, SRS3, and SRS4 were statistically similar to one another. These sites had the highest mean nitrogen measurements of all sites (Figure 4). The HSD test also determined that the mean nitrogen measurements of sites SRS1a, SRS1c, and SRS6 were statistically similar. These sites had the lowest mean nitrogen measurements (See Figure \@ref(fig:anova.nitro.site.plot)).

```{r, site analysis for nitrogen, include=FALSE}
#running anova to see if particular sites have more nitrogen
nitrogen.site.anova <- aov(data=FCE.site.nutrient.data, 
                           TN ~ SITE_NAME)
summary(nitrogen.site.anova)

#evaluating groups
nitrogen.site.anova.groups <- HSD.test(nitrogen.site.anova, 
                                       "SITE_NAME", 
                                       group = TRUE)
#printing groups
nitrogen.site.anova.groups

#results indicate that there are three groups of sites: a, b, and c.
```

```{r anova.nitro.site.plot, fig.cap= "Nitrogen measurements for mangrove habitat sites in the Florida Coastal Everglades", echo=FALSE, include=FALSE}
anova.nitro.site.plot <- ggplot(FCE.site.nutrient.data, aes(x=SITE_NAME, y=TN)) +
  geom_boxplot()+
  xlab(expression("Site Name")) +
  ylab(expression("Nitrogen (mcmol/L)")) +
  ggtitle(expression("Nitrogen Measurements for Sites"),
          subtitle="Florida Coastal Everglades") +
  annotate("text", x = 1, y = 1500, label = "bc") +
  annotate("text", x = 2, y = 1500, label = "bc") +
  annotate("text", x = 3, y = 1500, label = "a") +
  annotate("text", x = 4, y = 1500, label = "a") +
  annotate("text", x = 5, y = 1500, label = "a") +
  annotate("text", x = 6, y = 1500, label = "ab") +
  annotate("text", x = 7, y = 1500, label = "b") +
  annotate("text", x = 8, y = 1500, label = "c")
anova.nitro.site.plot
```

The results of the ANOVAs indicated that particular areas of Everglades National Park have elevated nutrient concentrations. Notably, sites SRS5 and SRS6 had the highest phosphorus concentrations, while sites SRS1d, SRS2, SRS3, and SRS4 have elevated nitrogen concentrations.

\newpage

# Summary and Conclusions

Overall, our results suggest that the nutrient conditions for mangroves in the Florida Everglades have varied both spatially and over time. Nutrient concentrations are indeed rising in the Park, and have increased substantially since the early 2000s, which may be cause for concern for nutrient-sensitive mangroves. It is apparent that some sites within the park have higher concentrations of nutrients than others. In particular, sites SRS5 and SRS6, which are located on the coast in the southwest corner of the park, have the highest phosphorus concentrations, and sites SRS1d, SRS2, and SRS3, which are located in the central and southwest areas of the park, have the highest nitrogen concentrations. This pattern may be observed due to the local hydrology, elevation, and slope of the sites. It is also possible that inputs from agriculture are greatest in particular areas of the park, due to the geographic location of the pollution source. Because agriculture is common in South Florida (FDACS 2022) and eutrophication can increase the mortality of mangroves (Lovelock et al. 2009), this may be a cause of concern for mangroves in the Everglades. We recommend conservation efforts in the park be focused to these identified, vulnerable sites. The Florida Coastal Everglades could also consider future research to determine the most resilient mangrove species to nitrogen and phosphorus inputs and focus re-seeding efforts in these conservation areas to those species. 

***paragraph to tie all together

\newpage

# References

Duke, N., Nagelkerken, I., Agardy, T., Wells, S., & Van Lavieren, H. (2014). The importance of mangroves to people: a call to action. United Nations Environment Programme World Conservation Monitoring Centre (UNEP-WCMC).

Florida Department of Agriculture and Consumer Services (FDACS (2022)). Florida Agriculture Overview and Statistics. https://www.fdacs.gov/Agriculture-Industry/Florida-Agriculture-Overview-and-Statistics 

Feller, I. C., Lovelock, C. E., & McKee, K. L. (2007). Nutrient addition differentially affects ecological processes of Avicennia germinans in nitrogen versus phosphorus limited mangrove ecosystems. Ecosystems, 10(3), 347-359.

Florida Coastal Everglades LTER. (2022). About. https://fcelter.fiu.edu/about/index.html. 

Lovelock, C. E., Ball, M. C., Martin, K. C., & C. Feller, I. (2009). Nutrient enrichment increases mortality of mangroves. PloS one, 4(5), e5600.

Nasser, Ramadan A.. “SPECIFIC GRAVITY , FIBER LENGTH AND CHEMICAL COMPONENTS OF CONOCARPUS ERECTUS AS AFFECTED BY TREE SPACING.” Journal of Agriculture & Environmental Science, Alexandria Univ., Egypt 7(3), (2008).

